---
title: "Regression Models Course Project"
author: "John Walker"
date: "18 Dec 2015"
output: rmarkdown::tufte_handout
---
# Overview

Looking a tooth growth data [E.W. Compton 1946](http://jn.nutrition.org/content/33/5/491.full.pdf) it provides a data frame with 60 observations on 3 variables:

1. len	 numeric Tooth length in microns.
2. supp	 factor	 Supplement type (VC or OJ).
3. dose	 numeric Dose in milligrams.

The course project is to perform basic exploratory analysis, use confidence intervals and/or hypothesis tests to compare tooth growth by supp and dose and to state conclusions from the tests and describe assumptions needed to form the conclusions.
```{r, echo=FALSE, warning=FALSE, message=FALSE}
require(ggplot2)
require(plyr)
require(xtable)
options(xtable.comment = FALSE)
options(xtable.booktabs = TRUE)

```


# Exploratory Data Analysis
An exploratory look at the tooth growth data set
```{r}
library(datasets)
data("ToothGrowth")
str(ToothGrowth)
xtable(head(ToothGrowth, 3), caption="First rows of ToothGrowth data")
summary(ToothGrowth)
```
shows 60 rows of the three variables - half for each supplement orange juice and vitamin C - grouped evenly into ten measurements at three dosage levels: 0.5mg, 1.0mg and 2.0 mg.
```{r}
table(ToothGrowth$supp, ToothGrowth$dose)
```

Looking at tooth growth length as a function of dosage for the two supplements it appears that orange juice promotes greater growth at the two lower dose levels and that growth is similar for the higher dose.
```{r, fig.cap="Exploratory point graph of ToothGrowth", warning=FALSE, message=FALSE}
qplot(dose, len, data=ToothGrowth, colour = supp, geom=c("point", "smooth"))
```

Calculating the mean, standard deviation and standard error of the mean for growth length over dosage groups by supplement confirms numerically the plot display: **mean values at lower dose levels are different between supplements and similar at the highest (2.0 mg) dose level**.

```{r}
ddply(ToothGrowth, c("dose", "supp"), summarise, mean = mean(len), sd = sd(len), sem=(sd(len)/sqrt(length(len))))
```

# Condifence Intervals: T-tests for mean difference by supplent type

```{r}
t.test(len ~ supp, var.equal=TRUE,  paired=FALSE, data=ToothGrowth)

```


# Interpretation

```{r}
t.test(len ~ supp, var.equal=FALSE, paired=FALSE, data=ToothGrowth)

```

# Assumptions
1. The Guinea pigs were assigned randomly to the groups of supplement and dosage.
2. These Guinea pigs represent the general Guinea pig population.
3. Variance of mean growth is equal across the groups.

# Conclusions
1. Across dosages the difference in tooth growth length between supplements is not significant.
2. The dosage amounts are statistically significant.
3. At the highest level 2.0 mg there is no difference in growth between supplements.
4. At levels 0.5mg and 1.0mg orange juice promotes greater tooth growth than the vitamin supplement.
